---
title: "Airport Network Analysis"
author: "Farhan"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: false
editor: visual
---

## Introduction

This project analyzes flight network data from March 2020 through October 2021, covering a critical period during the COVID-19 pandemic. The dataset includes comprehensive flight tracking information across multiple months.

## Dataset Overview

The analysis uses flight data spanning:

- **Time Period**: March 2020 - October 2021
- **Total Files**: 20 monthly datasets
- **Data Format**: CSV files with flight tracking information

### Data Collection Period

```{r}
#| echo: false
#| warning: false

library(lubridate)
library(dplyr)
library(ggplot2)

# Create a timeline of available data
months <- data.frame(
  date = seq(ymd("2020-03-01"), ymd("2021-10-01"), by = "month"),
  available = TRUE
)

ggplot(months, aes(x = date, y = 1)) +
  geom_point(size = 4, color = "#2E86AB") +
  geom_line(color = "#2E86AB", size = 1) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b %Y") +
  labs(title = "Flight Data Collection Timeline",
       subtitle = "March 2020 - October 2021",
       x = "Month",
       y = "") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Data Structure

Each monthly dataset contains flight information including:

- Flight identifiers
- Origin and destination airports
- Departure and arrival times
- Flight status information
- Aircraft details

## Project Goals

1. **Network Analysis**: Examine the structure and connectivity of the airport network
2. **Temporal Trends**: Analyze how flight patterns changed during the pandemic period
3. **Hub Identification**: Identify major airport hubs and their characteristics
4. **Route Analysis**: Study popular routes and their evolution over time

## Sample Data Analysis

```{r}
#| eval: false
#| echo: true

# Example: Loading and exploring a dataset
library(readr)
library(dplyr)

# Load a sample month
flights <- read_csv("dataset/flightlist_20210601_20210630.csv")

# Basic summary
summary(flights)

# Example analysis: Top airports by flight count
# top_airports <- flights %>%
#   count(origin_airport, sort = TRUE) %>%
#   head(10)
```

## Network Visualization

Future sections will include:

- Interactive network graphs showing airport connections
- Heat maps of flight density by time period
- Comparison of pre-pandemic vs pandemic flight patterns
- Geographic visualizations of routes

## Methodology

The analysis employs:

- **R** for data processing and statistical analysis
- **Network analysis** techniques to understand connectivity
- **Time series analysis** to identify trends
- **Geospatial analysis** for route visualization

## Next Steps

1. Load and combine all monthly datasets
2. Clean and standardize the data
3. Perform exploratory data analysis
4. Build network graphs
5. Analyze temporal patterns
6. Create interactive visualizations

## Data Source

Flight data collected from public flight tracking sources covering the period of March 2020 through October 2021.

---

*Last updated: `r Sys.Date()`*
